---
title: "TCGA mutational signature analysis in Section 5"
author: "DongHyuk Lee and Bin Zhu"
date: "December 24, 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### 1. Data and preprocessing
The signature loading data file <tt>TCGA_WES_sigProfiler_SBS_signatures_in_samples.csv</tt> can be downloaded from [here](https://www.synapse.org/#!Synapse:syn11801497) (available as of December 2020). 
It contains signature loadings of the extracted signatures (descried in Section 5 of the manuscript). 
Based on the COSMIC (Catalog Of Somatic Mutations In Cancer) database V3 ([here](https://cancer.sanger.ac.uk/cosmic/signatures/SBS/index.tt)), we removed signatures (e.g., SBS27) which were possibly caused by sequencing artefacts. 
Note that SBS84, SBS85, …, SBS90 in COSMIC database are not available in the above loading data (modified on January 29, 2020). 
In addition, we removed the duplicated tumor samples.
```{r}
TCGA_loadings1 <- read.csv("TCGA_WES_sigProfiler_SBS_signatures_in_samples.csv")
### Available signatures without possible sequencing artefacts (from https://cancer.sanger.ac.uk/cosmic/signatures/SBS/index.tt, where SBS84 through SBS90 are not available in the loadings.)
Sig_names <- c(paste0("SBS", 1:6), paste0("SBS7", letters[1:4]), paste0("SBS", 8:9),
           paste0("SBS10", letters[1:2]), paste0("SBS", 11:16), paste0("SBS17", letters[1:2]),
           paste0("SBS", 18:26), paste0("SBS", 28:42), "SBS44")
### Remove duplicated samples
sampleID <- substring(TCGA_loadings1$Sample.Names, 1, 12)
sum(duplicated(sampleID)) ## 38 duplicatesd
TCGA_loadings <- TCGA_loadings1[!duplicated(sampleID), c("Cancer.Types", Sig_names)]
head(TCGA_loadings)
table(TCGA_loadings$Cancer.Types)
```
In Section 5 of the manuscript, we focused on three gastrointestinal cancers: colorectal adenocarcinoma, esophageal adenocarcinoma and stomach adenocarcinoma. 
The following codes demonstrate the preprocessing steps to exclude signatures 1) when total loading counts are less than 10
and/or 2) when the number of subjects with non-zero loading counts is less than 5\% of the study size for each cancer type. 
We also included processed datasets, `Reduced_ColoRect-AdenoCa.csv` for colorectal adenocarcinoma, `Reduced_Eso-AdenoCa.csv`for esophageal adenocarcinoma, and `Reduced_Stomach-AdenoCa.csv` for stomach adenocarcinoma.
```{r}
### Extract gastrointestinal cancers
Gastro <- c("ColoRect-AdenoCa", "Eso-AdenoCa", "Stomach-AdenoCa")
### Total counts (Tcnt) and Nonzero counts (Ncnt) of signatures across cancer type 
Tcnt <- NULL
Ncnt <- NULL
for(typ in Gastro){
  Tcnt <- rbind(Tcnt, colSums(TCGA_loadings[TCGA_loadings$Cancer.Types == typ,-1]))
  Ncnt <- rbind(Ncnt, apply(TCGA_loadings[TCGA_loadings$Cancer.Types == typ,-1], 2, function(x) sum(x != 0)))
}
### Total loading counts per each cancer type
Tcnt.D <- cbind.data.frame(table(TCGA_loadings$Cancer.Types[TCGA_loadings$Cancer.Types %in% Gastro]), Tcnt)
Tcnt.D
### The number of subjects with non-zero loading counts
Ncnt.D <- cbind.data.frame(table(TCGA_loadings$Cancer.Types[TCGA_loadings$Cancer.Types %in% Gastro]), Ncnt)
Ncnt.D

### Extract columns (signatures) when total loading counts (Tcnt.D) > 10 and non-zero loading counts (Ncnt.D) > 0.05*(# of samples)
## colorectal adenocarcinoma
Sigs <- names(Tcnt.D)[-(1:2)][Tcnt.D[1, -(1:2)] > 10 & Ncnt.D[1, -(1:2)] > 0.05*Ncnt.D$Freq[1]]
Colo <- TCGA_loadings[TCGA_loadings$Cancer.Types == Gastro[1], Sigs]
head(Colo)
Colo_chk <- read.csv("Reduced_ColoRect-AdenoCa.csv")
#head(Colo_chk)
sum(abs(Colo_chk - Colo))

## esophageal adenocarcinoma
Sigs <- names(Tcnt.D)[-(1:2)][Tcnt.D[2, -(1:2)] > 10 & Ncnt.D[2, -(1:2)] > 0.05*Ncnt.D$Freq[2]]
Eso <- TCGA_loadings[TCGA_loadings$Cancer.Types == Gastro[2], Sigs]
head(Eso)
Eso_chk <- read.csv("Reduced_Eso-AdenoCa.csv")
#head(Eso_chk)
sum(abs(Eso_chk - Eso))

## stomach adenocarcinoma
Sigs <- names(Tcnt.D)[-(1:2)][Tcnt.D[3, -(1:2)] > 10 & Ncnt.D[3, -(1:2)] > 0.05*Ncnt.D$Freq[3]]
Stom <- TCGA_loadings[TCGA_loadings$Cancer.Types == Gastro[3], Sigs]
head(Stom)
Stom_chk <- read.csv("Reduced_Stomach-AdenoCa.csv")
#head(Stom_chk)
sum(abs(Stom_chk - Stom))
```
As a result, each column represents the signature loadings (as the number of point mutations contributed by a given signature) after preprocessing.
Eight signatures are included in colorectal adenocarcinoma while nine signatures are in other two cancer types. 
Among them, signatures 1, 2, 5 and 13 are present in all three cancer types but others are cancer-specific signatures.


### 2. Association tests
First we need to install `SKIT` package in the following way.
```{r}
### For Windows
# install.packages("SKIT_0.0.2.zip", repos = NULL)
### or
# install.packages("SKIT_0.0.2.tar.gz", repos = NULL, type = "source")
### For UNIX
# install.packages("SKIT_0.0.2.tar.gz", repos = NULL)
library(SKIT)
```
Here, we present results for one cancer type, colorectal adenocarcinoma, but results corresponding to other cancer types can be obtained similarly as described in Section 2.1.
```{r}
### All possible combinations of signatures of colorectal adenocarcinoma
Allpairs <- combn(names(Colo), 2)
ncol(Allpairs)
### Normalization with respect to total counts (Section 2.2 and Section 5 in the manuscript)
Colo_norm <- t(apply(Colo,1, function(x){x/sum(x)}))
Colo_norm[is.nan(Colo_norm)] <- 0
```
In the manuscript, we used $B = 30,000$ bootstrap samples to distinguish small scale $p$-values.
Since the tests of all possible combinations (28 combinations) are time consuming, we show the result of one signature pair (signature 2 and signature 17b) as an example. 
The entire results are provided in <tt>online_supplements.zip</tt> file. 
To replicate the same results, please use `set.seed(1)` for all combinations regardless of cancer types.
```{r}
Allpairs[,11]
tmp <- Colo_norm[, Allpairs[,11]]
x <- tmp[,1]
y <- tmp[,2]
set.seed(1)
system.time( re <- skit(x, y, nboot = 30000) )
re
```
The computation is conducted in a Dell Latitude 7490 laptop with Intel(R) Core(TM) i7-8650 CPU @ 1.90GHz.
We would like to point out that computing time may vary depending on the sample size as well as the number of non-zero elements.
`skit()` function returns 1) `re$tests` which contains 5 test statistics (the overall SKIT test statistic (`T`) and its components (`T1`, `T2`, `T3`, `T4`));
2) `re$pvalues` p-values corresponding to test statistics in `re$tests`;
and 3) `re$bandwidth.obs` the bandwidth of kernel density (see Section 3 for details).

We have conducted association tests for all possible pairs and <tt>SKIT_Result_ColoRect-AdenoCa.csv</tt> contains test results of all 28 combinations for colorectal adenocarcinoma.
```{r}
Colo_entire <- read.csv("SKIT_Result_ColoRect-AdenoCa.csv", stringsAsFactors = FALSE)
Colo_entire[10:14,]#[Colo_entire$Sig1 == "SBS2" & Colo_entire$Sig2 == "SBS17b",]
```
Each row of `Colo_entire` shows `skit()` result for a pair of signatures. For example, the 11th row presents the results for the pair (SBS2, SBS17b).
The entire results can be obtained via the following code, but we commented out due to long computation time.
```{r}
# Colo_re <- NULL
# for(i in 1:ncol(Allpairs)){
#   tmp <- Colo_norm[, Allpairs[,i]]
#   x <- tmp[,1]
#   y <- tmp[,2]
#   set.seed(1)
#   re <- skit(x, y, nboot = 30000)
#   tmp1 <- matrix(c(re$bandwidth.obs, re$tests, re$pvalues), nrow = 1)
#   colnames(tmp1) <- c("bws", "T", paste0("T", 1:4), "p", paste0("p", 1:4))
#   Colo_re <- rbind(Colo_re, tmp1)
# }
# Colo_re
```
#### 2.1 Test results for other cancer type
`skit()` results of the signature pair 2 and 17b from esophageal and stomach adenocarcinoma, can be obtained from the following codes.
<tt>SKIT_Result_Eso-AdenoCa.csv</tt> and <tt>SKIT_Result_Stomach-AdenoCa.csv</tt> contain test results of all combinations.
```{r}
Eso_norm <- t(apply(Eso,1, function(x){x/sum(x)}))
Eso_norm[is.nan(Eso_norm)] <- 0
Allpairs <- combn(names(Eso), 2)
tmp <- Eso_norm[, Allpairs[,14]]
x <- tmp[,1]
y <- tmp[,2]
set.seed(1)
re <- skit(x, y, nboot = 30000)
Eso_entire <- read.csv("SKIT_Result_Eso-AdenoCa.csv", stringsAsFactors = FALSE)
re
Eso_entire[14,]
```

```{r}
Stom_norm <- t(apply(Stom,1, function(x){x/sum(x)}))
Stom_norm[is.nan(Stom_norm)] <- 0
Allpairs <- combn(names(Stom_chk), 2)
tmp <- Stom_norm[, Allpairs[,13]]
x <- tmp[,1]
y <- tmp[,2]
set.seed(1)
re <- skit(x, y, nboot = 30000)
Stom_entire <- read.csv("SKIT_Result_Stomach-AdenoCa.csv", stringsAsFactors = FALSE)
re
Stom_entire[13,]
```



### 3. Creating Table 2
The following codes are used to generate Table 2.
```{r}
Eso_entire <- read.csv("SKIT_Result_Eso-AdenoCa.csv", stringsAsFactors = FALSE)
Stom_entire <- read.csv("SKIT_Result_Stomach-AdenoCa.csv", stringsAsFactors = FALSE)
Skit_Colo_FDR <- cbind(Colo_entire[,1:2], round(Colo_entire[,-(1:3)],4),
                       BH = round(p.adjust(Colo_entire$p, method = "BH"), 4),
                       BH1 = round(p.adjust(Colo_entire$p1, method = "BH"), 4),
                       BH2 = round(p.adjust(Colo_entire$p2, method = "BH"), 4),
                       BH3 = round(p.adjust(Colo_entire$p3, method = "BH"), 4),
                       BH4 = round(p.adjust(Colo_entire$p4, method = "BH"), 4))
Skit_Eso_FDR <- cbind(Eso_entire[,1:2], round(Eso_entire[,-(1:3)],4),
                       BH = round(p.adjust(Eso_entire$p, method = "BH"), 4),
                       BH1 = round(p.adjust(Eso_entire$p1, method = "BH"), 4),
                       BH2 = round(p.adjust(Eso_entire$p2, method = "BH"), 4),
                       BH3 = round(p.adjust(Eso_entire$p3, method = "BH"), 4),
                       BH4 = round(p.adjust(Eso_entire$p4, method = "BH"), 4))
Skit_Stom_FDR <- cbind(Stom_entire[,1:2], round(Stom_entire[,-(1:3)],4),
                      BH = round(p.adjust(Stom_entire$p, method = "BH"), 4),
                      BH1 = round(p.adjust(Stom_entire$p1, method = "BH"), 4),
                      BH2 = round(p.adjust(Stom_entire$p2, method = "BH"), 4),
                      BH3 = round(p.adjust(Stom_entire$p3, method = "BH"), 4),
                      BH4 = round(p.adjust(Stom_entire$p4, method = "BH"), 4))
Tab2 <- rbind(
          Skit_Colo_FDR[Skit_Colo_FDR$Sig1 %in% c("SBS2", "SBS13") &
                        Skit_Colo_FDR$Sig2 %in% c("SBS17a", "SBS17b"),
                        c("Sig1", "Sig2", "BH", "BH1", "BH2", "BH3", "BH4")],
          Skit_Eso_FDR[Skit_Eso_FDR$Sig1 %in% c("SBS2", "SBS13") &
                         Skit_Eso_FDR$Sig2 %in% c("SBS17a", "SBS17b"),
                        c("Sig1", "Sig2", "BH", "BH1", "BH2", "BH3", "BH4")],
          Skit_Stom_FDR[Skit_Stom_FDR$Sig1 %in% c("SBS2", "SBS13") &
                        Skit_Stom_FDR$Sig2 %in% c("SBS17a", "SBS17b"),
                        c("Sig1", "Sig2", "BH", "BH1", "BH2", "BH3", "BH4")]
        )
names(Tab2) <- c("Sig1", "Sig2", "T", "T1", "T2", "T3", "T4")
row.names(Tab2) <- NULL
Tab2 <- cbind(Type = c("colorectal", "", "esophageal", "", "", "", "stomach", "", "", ""), Tab2)
library(knitr)
kable(Tab2, caption = "Table 2. FDR adjusted p-values of all active pairs between 17a/b and
      APOBEC-mediated signatures for gastrointestinal cancers.")
```


### 4. Generating Figure S.3
For the signature pairs 2 and 17b, we provide the following scatterplots to investigate in detail.
```{r, fig.height=4,fig.width=12}
library(ggplot2)
library(gridExtra)
# Colorectal adenocarcinoma 2 and 17b
x <- Colo_norm[, "SBS2"]
y <- Colo_norm[, "SBS17b"]
Colo_D <- data.frame(x,y)
Colo_D$idx <- factor(ifelse(Colo_D$x != 0 & Colo_D$y != 0, 1, 0))
Colo217b <- ggplot(data = Colo_D, aes(x = x, y = y, shape = idx)) + geom_point(size = 2) +
    xlab("Signature 2") + ylab("Signature 17b") +
    scale_shape_manual(values = c(19, 1)) + ggtitle("(a) Colorectal adenocarcinoma") +
    theme(legend.position = "none", panel.background = element_blank(),
          axis.line = element_blank(), axis.ticks = element_blank(),
          axis.title = element_text(size = 10), axis.text = element_text(size = 10),
          plot.title = element_text(size = 10, hjust = 0)) +
    geom_hline(yintercept = 0) + geom_vline(xintercept = 0) +
    scale_x_continuous(limits = c(0, 0.55), expand = c(0, 0.01)) +
    scale_y_continuous(limits = c(0, 0.55), expand = c(0, 0.01))
# Esophageal adenocarcinoma 2 and 17b
Eso_norm <- t(apply(Eso,1, function(x){x/sum(x)}))
Eso_norm[is.nan(Eso_norm)] <- 0
x <- Eso_norm[, "SBS2"]
y <- Eso_norm[, "SBS17b"]
Eso_D <- data.frame(x,y)
Eso_D$idx <- factor(ifelse(Eso_D$x != 0 & Eso_D$y != 0, 1, 0))
Eso217b <- ggplot(data = Eso_D, aes(x = x, y = y, shape = idx)) + geom_point(size = 2) +
    xlab("Signature 2") + ylab("Signature 17b") +
    scale_shape_manual(values = c(19, 1)) + ggtitle("(b) Esophageal adenocarcinoma") +
    theme(legend.position = "none", panel.background = element_blank(),
          axis.line = element_blank(), axis.ticks = element_blank(),
          axis.title = element_text(size = 10), axis.text = element_text(size = 10),
          plot.title = element_text(size = 10, hjust = 0)) +
    geom_hline(yintercept = 0) + geom_vline(xintercept = 0) +
    scale_x_continuous(limits = c(0, 0.55), expand = c(0, 0.01)) +
    scale_y_continuous(limits = c(0, 0.55), expand = c(0, 0.01))
# Stomach adenocarcinoma 2 and 17b
Stom_norm <- t(apply(Stom,1, function(x){x/sum(x)}))
Stom_norm[is.nan(Stom_norm)] <- 0
x <- Stom_norm[, "SBS2"]
y <- Stom_norm[, "SBS17b"]
Stom_D <- data.frame(x,y)
Stom_D$idx <- factor(ifelse(Stom_D$x != 0 & Stom_D$y != 0, 1, 0))
Stom217b <- ggplot(data = Stom_D, aes(x = x, y = y, shape = idx)) + geom_point(size = 2) +
    xlab("Signature 2") + ylab("Signature 17b") +
    scale_shape_manual(values = c(19, 1)) + ggtitle("(c) Stomach adenocarcinoma") +
    theme(legend.position = "none", panel.background = element_blank(),
          axis.line = element_blank(), axis.ticks = element_blank(),
          axis.title = element_text(size = 10), axis.text = element_text(size = 10),
          plot.title = element_text(size = 10, hjust = 0)) +
    geom_hline(yintercept = 0) + geom_vline(xintercept = 0) +
    scale_x_continuous(limits = c(0, 0.55), expand = c(0, 0.01)) +
    scale_y_continuous(limits = c(0, 0.55), expand = c(0, 0.01))
grid.arrange(Colo217b, Eso217b, Stom217b, nrow = 1)
```


### 5. Making Figure 4
The following plots are modification from the one found in http://jokergoo.github.io/blog/html/large_matrix_circular.html.
The source code of function `cir.corr()` can be found in "cir.corr.R" but please note that the `circlize` package is required.
```{r, fig.height = 12, fig.width = 12}
source("cir.corr.R")
library(circlize)
par(mfrow = c(2,2))
cir.corr("Colo")
cir.corr("Eso")
cir.corr("Stom")
```


### 6. Session Info
```{r}
sessionInfo()
```